---
title: "Citation Audit Workflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Citation Audit Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = TRUE,
  warning  = FALSE
)
```

```{r css, echo = FALSE, results = "asis"}
# Extra CSS: give vignette tables breathing room
cat("
<style>
table { font-size: 13px; }
th { background: #f0f0f0; color: #333; padding: 6px 10px; border-bottom: 2px solid #ccc; }
td { padding: 5px 10px; vertical-align: top; }
.status-auto      { background: #FFF2CC; }
.status-yes       { background: #D9EAD3; }
.status-no        { background: #F4CCCC; }
.status-corrected { background: #FCE5CD; }
.status-no_match  { background: #E2EFDA; }
.status-na        { background: #f5f5f5; color: #999; }
pre { font-size: 12px; }
</style>
")
```

## The problem

When a report is drafted with LLM assistance, a specific failure mode arises: the citation key is real, the prose sounds authoritative, but the cited source does not actually support the specific claim made. A percentage gets transposed, a finding from one paper gets attributed to another, or an experimental result gets quietly generalised.

`cred` addresses this systematically. For every `@citekey` in a bookdown report it locates the source document, finds the best-matching passage, and surfaces it alongside the original claim so a human can make the call.

## Workflow at a glance

```
Rmd files  ──►  audit CSV  ──►  Zotero lookup  ──►  auto-fill quotes
                                                           │
                                              ┌────────────┴─────────────┐
                                         auto-match                  no match
                                         (review)                 (manual check)
                                              │
                                         Review app  ──►  yes / no / corrected
```

| Step | Function | What it does |
|------|----------|-------------|
| 1 | `crd_aud_write()` | Scan Rmd files → extract citations with paraphrase context → write CSV scaffold |
| 2 | `crd_aud_verify_all()` | Look up Zotero attachments → search each source → fill `quote` and `verified` |
| 3 | `crd_aud_sort()` | Sort `auto` rows first so reviewable rows are at the top |
| 4 | `crd_aud_summary()` | Print status snapshot and PDF attachment priority list |
| 5 | `crd_aud_review()` | Launch interactive Shiny app — paraphrase vs quote side by side, dropdown to set `verified` |

---

## Walkthrough with toy data

`cred` ships realistic toy source documents and a minimal Zotero SQLite in `inst/extdata/`:

```
inst/extdata/
├── zotero.sqlite              # 3 items: smith2020, jones2019, doe2021 (no attachment)
└── storage/
    ├── TOYPDF1/salmon_habitat.pdf      # Smith et al. 2020 — 4 paragraphs
    └── TOYDOC1/beaver_ecology.docx     # Jones 2019 — 4 paragraphs
```

```{r paths}
library(cred)

toy_zotero_dir <- system.file("extdata", package = "cred")
```

---

### Step 1 — Write the audit CSV

`crd_aud_write()` scans all chapter Rmd files (names beginning with four digits), extracts every `@citekey` with the surrounding sentence as `paraphrase`, and writes a CSV scaffold with blank `verified`, `quote`, and `notes` columns.

The toy Rmd files reflect three common paraphrase patterns:

1. **Numeric claims** — statistics copied closely from the source ("62% lower", "38% were barriers"). These score highest because numbers are the most discriminative tokens.
2. **Qualitative paraphrases** — same finding, different words, no numbers. These test whether token overlap still finds the right passage when an author rewrites rather than transcribes.
3. **Wrong-paper citation** — a Pacific Salmon Treaty harvest claim cites `smith2020SalmonHabitat`, a freshwater habitat quality paper. It has no content about harvest allocation or treaty negotiations, so no passage scores above the threshold → `verified = "no_match"`. This is the failure mode `cred` is designed to surface: the citation key is real, the prose sounds plausible, but the source does not support the claim.

Each sentence is on a single line so `crd_aud_write()` captures the full paraphrase.

```{r write-audit}
rmd_dir    <- tempfile("rmd_")
audit_file <- tempfile("audit_", fileext = ".csv")
dir.create(rmd_dir)

# Chapter 1 — habitat quality: 5 claims, 3 citation keys
# smith2020 has a PDF attachment; doe2021 has metadata only (no PDF).
# Mix of numeric claims (rows 1-3), a qualitative paraphrase (row 4),
# an unsupported claim that produces no_match (row 5), and an NA row (row 6).
writeLines(c(
  "# Habitat Quality {#habitat}",
  "",
  "Reaches where spawning substrate embeddedness exceeded 25% had egg-to-fry survival rates 62% lower than adjacent low-embeddedness sites, making fine sediment the dominant limiting factor for chinook reproduction [@smith2020SalmonHabitat].",
  "",
  "Removal of riparian conifers within 30 m of the bankfull channel raised daily maximum summer stream temperatures by an average of 4.2 degrees Celsius compared to unlogged reference reaches, exceeding the upper thermal tolerance of juvenile chinook during late-summer rearing [@smith2020SalmonHabitat].",
  "",
  "Fish passage assessments at 1,200 road-stream crossings found that 38% were rated as full or partial barriers to adult chinook migration, primarily due to culverts with outlet drops exceeding 15 cm [@smith2020SalmonHabitat].",
  "",
  "Loss of streamside forest cover eliminates the shade that keeps summer water temperatures within the thermal tolerance of juvenile chinook, degrading rearing conditions in harvested riparian zones [@smith2020SalmonHabitat].",
  "",
  "The Pacific Salmon Treaty allocates harvest between Canada and the United States based on abundance indicators and conservation thresholds negotiated through the Pacific Salmon Commission [@smith2020SalmonHabitat].",
  "",
  "Stream temperature governs the spatial distribution of juvenile salmon and limits rearing area during summer low flows [@doe2021NoFile]."
), file.path(rmd_dir, "0100-habitat.Rmd"))

# Chapter 2 — beaver ecology: 5 claims, all jones2019 (docx attachment)
# Row 5 is a qualitative paraphrase of the floodplain side-channel paragraph.
writeLines(c(
  "# Beaver Ecology {#beaver}",
  "",
  "Electrofishing surveys recorded juvenile salmonid densities 3.4 times higher in beaver pond habitat than in adjacent free-flowing reaches, with slow-water refuges providing shelter from winter spates while maintaining access to benthic invertebrate prey [@jones2019BeaverEcology].",
  "",
  "Experimental reintroduction of beavers to an incised stream reach reduced peak daily flows by 31% during late-season storm events and extended summer baseflow duration by five weeks due to increased groundwater storage in pond sediments [@jones2019BeaverEcology].",
  "",
  "Archival records indicate that beaver pond habitat declined by approximately 60% between 1850 and 1950 due to commercial trapping pressure, reducing overwinter rearing capacity for juvenile salmonids [@jones2019BeaverEcology].",
  "",
  "Juvenile chinook tagged in the main channel were detected in floodplain side channels during high-flow events, where they remained for 12 to 34 days and achieved growth rates 40% higher than fish remaining in the mainstem [@jones2019BeaverEcology].",
  "",
  "Beaver-mediated floodplain connectivity allows juvenile chinook to access off-channel rearing areas during high flows, where growth rates substantially exceed those of fish confined to the mainstem channel [@jones2019BeaverEcology]."
), file.path(rmd_dir, "0200-beaver.Rmd"))

crd_aud_write(rmd_dir = rmd_dir, out_file = audit_file)
```

```{r show-scaffold, echo = FALSE}
d <- readr::read_csv(audit_file, show_col_types = FALSE)
DT::datatable(
  d[, c("section", "citation_key", "paraphrase", "verified")],
  rownames = FALSE,
  options  = list(
    pageLength = 5, scrollX = TRUE, dom = "tip",
    columnDefs = list(
      list(width = "90px",  targets = 0),
      list(width = "180px", targets = 1),
      list(width = "380px", targets = 2),
      list(width = "70px",  targets = 3)
    )
  ),
  caption = "Audit scaffold — 11 rows, verified column blank."
)
```

Eleven rows across two chapters — five per source plus the `doe2021NoFile` row. The `verified` column is blank — no matching has happened yet.

---

### Step 2 — Look up Zotero source paths

`crd_zot_src_lookup()` queries the Zotero SQLite database (read-only, immutable URI — safe to run while Zotero is open) and returns file paths for any attached PDFs or Word documents.

```{r lookup}
keys    <- unique(d$citation_key)
sources <- crd_zot_src_lookup(keys, zotero_dir = toy_zotero_dir)
sources[, c("citation_key", "src_type")]
```

`doe2021NoFile` has no attachment — it was added to Zotero by DOI lookup (metadata only, no PDF downloaded). The warning is expected; the row will remain `NA` in the audit. `crd_aud_summary()` will flag it as a priority to attach.

---

### Step 3 — Auto-fill quotes

`crd_aud_verify_all()` loads each source document once, then for every unverified row matching that key it scores paragraphs/pages against the `paraphrase` using token overlap. It fills:

- **`quote`** — the best-matching passage from the source
- **`page_or_section`** — paragraph index (docx) or page number (pdf)
- **`verified`** — `"auto"` if score ≥ `min_score`; `"no_match"` if below threshold

```{r verify}
crd_aud_verify_all(
  audit_file = audit_file,
  sources    = sources
)
```

```{r show-verified, echo = FALSE}
d2 <- readr::read_csv(audit_file, show_col_types = FALSE)

# Colour verified column for the table
status_col <- ifelse(
  is.na(d2$verified), "<span style='color:#bbb'>NA</span>",
  dplyr::case_when(
    d2$verified == "auto"     ~ "<span style='background:#FFF2CC;padding:2px 6px;border-radius:3px'>auto</span>",
    d2$verified == "no_match" ~ "<span style='background:#E2EFDA;padding:2px 6px;border-radius:3px'>no_match</span>",
    TRUE ~ d2$verified
  )
)

d2_show <- data.frame(
  citation_key    = d2$citation_key,
  paraphrase      = d2$paraphrase,
  quote           = ifelse(is.na(d2$quote), "", d2$quote),
  loc             = ifelse(is.na(d2$page_or_section), "", d2$page_or_section),
  verified        = status_col,
  stringsAsFactors = FALSE
)

DT::datatable(
  d2_show,
  rownames = FALSE,
  escape   = FALSE,
  colnames = c("citation_key", "paraphrase", "quote", "p/§", "verified"),
  options  = list(
    pageLength = 5, scrollX = TRUE, dom = "tip",
    columnDefs = list(
      list(width = "160px", targets = 0),
      list(width = "270px", targets = 1),
      list(width = "270px", targets = 2),
      list(width = "40px",  targets = 3),
      list(width = "80px",  targets = 4)
    )
  ),
  caption = "After crd_aud_verify_all(): auto = matched, no_match = below threshold, NA = no source file."
)
```

**Reading the outcomes:**

- **`auto`** (yellow) — a passage was found in the source scoring above the threshold. The `quote` column shows what the source actually says. Your job: read both, decide if they match.
- **`no_match`** (light green) — the source file exists, but no paragraph scored above the threshold. This happens when the report's phrasing is too different from the source text. It needs manual verification.
- **`NA`** (grey) — no source file is attached in Zotero. Nothing to match against.

Notice the Pacific Salmon Treaty row: it cites `smith2020SalmonHabitat`, a freshwater habitat quality paper, for a claim about international harvest allocation. No passage in the source covers treaty negotiations — `no_match` flags it immediately. This is exactly the kind of mismatch `cred` is designed to surface: the citation key is real, the prose sounds credible, but the source simply does not support the specific claim.

---

### Step 4 — Sort for review

`crd_aud_sort()` rearranges rows so the most actionable are at the top. The `"status"` order is:

`auto` → `no_match` → `corrected` → `no` → `context` → `yes` → `NA`

```{r sort}
crd_aud_sort(audit_file, by = "status")
```

`sort_index` is assigned at write time and never changes, so `crd_aud_sort(by = "report")` always restores the original chapter order.

---

### Step 5 — Summary

`crd_aud_summary()` prints a live status snapshot and — crucially — tells you which PDFs to attach in Zotero to unlock the most NA rows.

```{r summary}
crd_aud_summary(audit_file)
```

In a real report with 200+ citations, the NA sources table is your action list: attach the top-ranked PDF, re-run `crd_aud_verify_all()`, and unlock several rows at once.

---

### Step 6 — Interactive review

`crd_aud_review()` launches a local Shiny app in your browser. It is the recommended interface for working through `auto` rows.

```r
crd_aud_review("background/citation_audit.csv")
```

The app opens filtered to `auto` rows. For each row:

1. Click the row — full `paraphrase` and `quote` appear side by side below the table
2. Read both — does the source support the claim?
3. Set `verified` in the dropdown: `yes`, `no`, or `corrected`
4. Add a `notes` entry if the claim needs correction or context
5. Click **Update row**, move to the next
6. Click **Save to CSV** when done — the plain-text CSV is updated, git history preserved

The app also shows a live progress counter: *N reviewed · N auto · N total*.

---

## The three outcomes explained

| Outcome | What it means | What to do |
|---------|--------------|-----------|
| `auto` | Source file found; passage scored above threshold | Open review app, read quote vs paraphrase, set yes/no/corrected |
| `no_match` | Source file found but no paragraph matched | Check source manually; if claim is valid, set `context` or `yes`; if wrong, set `corrected` |
| `NA` | No source file attached in Zotero | Attach PDF in Zotero → re-run `crd_aud_verify_all()` |

**`no_match` is not an error.** It occurs when:

- The report's phrasing diverges substantially from the source (a sign the claim may have been paraphrased beyond what the source says — worth checking)
- The source provides context rather than a direct quotable finding
- Multiple sources are cited together (`[@a; @b]`) and the paraphrase blends both

---

## Claim type screening

`crd_aud_scr_risk()` populates the `claim_type` column to prioritise your review effort. Statistics (numbers, percentages, years) carry the highest hallucination risk; context citations the lowest.

```{r risk}
d3 <- readr::read_csv(audit_file, show_col_types = FALSE)
d3 <- crd_aud_scr_risk(d3)
table(d3$claim_type)
```

For domain-specific terms, pass an extra pattern:

```r
# Flag species names as statistics for a salmonid report
d3 <- crd_aud_scr_risk(d3, statistic_extra = "chinook|sockeye|coho|pink|chum")
```

Review `statistic` rows first.

---

## Working with your real report

```r
library(cred)

# ── One-time setup ──────────────────────────────────────────────────────────
# Generate the audit CSV from project root (chapter Rmd files only)
crd_aud_write(rmd_dir = ".", out_file = "background/citation_audit.csv")

# ── After adding new Rmd content ────────────────────────────────────────────
# Merge new citations without losing manual edits
crd_aud_upd("background/citation_audit.csv")

# ── Auto-fill from Zotero ───────────────────────────────────────────────────
# Zotero must be open for BBT citation keys; not needed for the SQLite lookup
crd_aud_verify_all("background/citation_audit.csv")

# ── Check progress ──────────────────────────────────────────────────────────
crd_aud_summary("background/citation_audit.csv")

# ── Review session ──────────────────────────────────────────────────────────
# Sort auto rows to the top, open the review app
crd_aud_sort("background/citation_audit.csv", by = "status")
crd_aud_review("background/citation_audit.csv")

# ── Restore report order when done ─────────────────────────────────────────
crd_aud_sort("background/citation_audit.csv", by = "report")
```

### Attaching missing PDFs

`crd_aud_summary()` lists NA sources ranked by claim count — one attachment can unlock many rows. In Zotero:

1. Find the item (use the `citation_key` to search)
2. Right-click → **Add Attachment → Attach File**
3. Re-run `crd_aud_verify_all()` — newly attached files are picked up automatically

### How source matching works

The token scoring approach is simple and fast:

1. Strip R inline expressions (`` `r ...` ``), `@citekey` markers, and markdown punctuation from the `paraphrase`
2. Extract word tokens (≥ 4 characters) and numeric tokens (digit sequences, percentages) separately — numbers are the most discriminative signals for factual claims
3. Score each paragraph/page: proportion of query tokens found in the passage
4. Accept the top-scoring passage if score ≥ `min_score` (default 0.2)

**What 0.2 means:** 1 in 5 query tokens must appear in the candidate paragraph. With a 15-token query, 3 tokens need to match. This is deliberately permissive — a false positive (wrong paragraph shown as `auto`) costs you a few seconds of review; a false negative (right paragraph called `no_match`) means you search the PDF yourself.

**`min_score` is a parameter in every matching function:**

```r
crd_aud_verify_all("citation_audit.csv", min_score = 0.2)   # default
crd_pdf_srch_clm(pdf_txt, paraphrase, min_score = 0.2)
crd_docx_srch_clm(docx_txt, paraphrase, min_score = 0.2)
```

| Situation | Suggested `min_score` |
|---|---|
| Short or generic paraphrases — many false positives | 0.3 – 0.4 |
| Default — specific factual claims with some numbers | 0.2 |
| Long dense sources where good matches score low | 0.1 – 0.15 |

After adjusting, re-run `crd_aud_verify_all(..., overwrite_verified = TRUE)` to reprocess all machine-assigned rows without touching human-reviewed ones (`yes`, `no`, `corrected`, `context`).
