---
title: "Citation Audit Workflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Citation Audit Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = TRUE,
  warning  = FALSE
)
```

```{r css, echo = FALSE, results = "asis"}
# Extra CSS: give vignette tables breathing room
cat("
<style>
table { font-size: 13px; }
th { background: #1a1a1a; color: white; padding: 6px 10px; }
td { padding: 5px 10px; vertical-align: top; }
.status-auto      { background: #FFF2CC; }
.status-yes       { background: #D9EAD3; }
.status-no        { background: #F4CCCC; }
.status-corrected { background: #FCE5CD; }
.status-no_match  { background: #E2EFDA; }
.status-na        { background: #f5f5f5; color: #999; }
pre { font-size: 12px; }
</style>
")
```

## The problem

When a report is drafted with LLM assistance, a specific failure mode arises: the citation key is real, the prose sounds authoritative, but the cited source does not actually support the specific claim made. A percentage gets transposed, a finding from one paper gets attributed to another, or an experimental result gets quietly generalised.

`cred` addresses this systematically. For every `@citekey` in a bookdown report it locates the source document, finds the best-matching passage, and surfaces it alongside the original claim so a human can make the call.

## Workflow at a glance

```
Rmd files  ──►  audit CSV  ──►  Zotero lookup  ──►  auto-fill quotes
                                                           │
                                              ┌────────────┴─────────────┐
                                         auto-match                  no match
                                         (review)                 (manual check)
                                              │
                                         Review app  ──►  yes / no / corrected
```

| Step | Function | What it does |
|------|----------|-------------|
| 1 | `crd_aud_write()` | Scan Rmd files → extract citations with paraphrase context → write CSV scaffold |
| 2 | `crd_aud_verify_all()` | Look up Zotero attachments → search each source → fill `quote` and `verified` |
| 3 | `crd_aud_sort()` | Sort `auto` rows first so reviewable rows are at the top |
| 4 | `crd_aud_summary()` | Print status snapshot and PDF attachment priority list |
| 5 | `crd_aud_review()` | Launch interactive Shiny app — paraphrase vs quote side by side, dropdown to set `verified` |

---

## Walkthrough with toy data

`cred` ships realistic toy source documents and a minimal Zotero SQLite in `inst/extdata/`:

```
inst/extdata/
├── zotero.sqlite              # 3 items: smith2020, jones2019, doe2021 (no attachment)
└── storage/
    ├── TOYPDF1/salmon_habitat.pdf      # Smith et al. 2020 — 4 paragraphs
    └── TOYDOC1/beaver_ecology.docx     # Jones 2019 — 4 paragraphs
```

```{r paths}
library(cred)

toy_zotero_dir <- system.file("extdata", package = "cred")
```

---

### Step 1 — Write the audit CSV

`crd_aud_write()` scans all chapter Rmd files (names beginning with four digits), extracts every `@citekey` with the surrounding sentence as `paraphrase`, and writes a CSV scaffold with blank `verified`, `quote`, and `notes` columns.

The toy Rmd files below reflect common patterns in scientific reports: specific statistics, methodological findings, and one claim the source does not support (the floodplain growth claim in the habitat chapter is not covered by `smith2020SalmonHabitat` — its paragraph 4 covers large woody debris, not floodplain rearing). This is deliberate: we want to see the `no_match` outcome.

```{r write-audit}
rmd_dir    <- tempfile("rmd_")
audit_file <- tempfile("audit_", fileext = ".csv")
dir.create(rmd_dir)

# Chapter 1 — habitat quality: 4 claims, 3 citation keys
# smith2020 has a PDF attachment; doe2021 has metadata only (no PDF)
writeLines(c(
  "# Habitat Quality {#habitat}",
  "",
  "Spawning substrate embeddedness exceeding 25% reduces egg-to-fry",
  "survival by more than half compared to low-embeddedness sites",
  "[@smith2020SalmonHabitat].",
  "",
  "Removal of riparian conifers within 30 m of the channel raises",
  "maximum summer stream temperatures by 3 to 7 degrees Celsius",
  "[@smith2020SalmonHabitat].",
  "",
  "Of 1,200 road crossings assessed in the watershed, 38% were rated",
  "as full or partial barriers to adult chinook migration",
  "[@smith2020SalmonHabitat].",
  "",
  "Floodplain rearing habitat supports juvenile growth rates substantially",
  "higher than mainstem conditions [@smith2020SalmonHabitat].",
  "",
  "Stream temperature governs the spatial distribution of juvenile salmon",
  "and limits rearing area during summer low flows [@doe2021NoFile]."
), file.path(rmd_dir, "0100-habitat.Rmd"))

# Chapter 2 — beaver ecology: 4 claims, all jones2019 (docx attachment)
writeLines(c(
  "# Beaver Ecology {#beaver}",
  "",
  "Beaver ponds support juvenile salmonid densities more than three times",
  "higher than adjacent free-flowing reaches during winter",
  "[@jones2019BeaverEcology].",
  "",
  "Beaver reintroduction reduced peak flows by approximately 30% and",
  "extended summer baseflow duration by five weeks in experimental reaches",
  "[@jones2019BeaverEcology].",
  "",
  "Commercial trapping reduced beaver pond habitat by approximately 60%",
  "between 1850 and 1950, with lasting consequences for overwinter",
  "juvenile survival [@jones2019BeaverEcology].",
  "",
  "Juvenile chinook using floodplain side channels during high-flow events",
  "grew 40% faster than fish remaining in the mainstem",
  "[@jones2019BeaverEcology]."
), file.path(rmd_dir, "0200-beaver.Rmd"))

crd_aud_write(rmd_dir = rmd_dir, out_file = audit_file)
```

```{r show-scaffold, echo = FALSE}
d <- readr::read_csv(audit_file, show_col_types = FALSE)
DT::datatable(
  d[, c("section", "citation_key", "paraphrase", "verified")],
  rownames = FALSE,
  options  = list(
    pageLength = 9, scrollX = TRUE, dom = "t",
    columnDefs = list(
      list(width = "90px",  targets = 0),
      list(width = "180px", targets = 1),
      list(width = "380px", targets = 2),
      list(width = "70px",  targets = 3)
    )
  ),
  caption = "Audit scaffold — 9 rows, verified column blank."
)
```

Nine rows across two chapters. The `verified` column is blank — no matching has happened yet.

---

### Step 2 — Look up Zotero source paths

`crd_zot_src_lookup()` queries the Zotero SQLite database (read-only, immutable URI — safe to run while Zotero is open) and returns file paths for any attached PDFs or Word documents.

```{r lookup}
keys    <- unique(d$citation_key)
sources <- crd_zot_src_lookup(keys, zotero_dir = toy_zotero_dir)
sources[, c("citation_key", "src_type")]
```

`doe2021NoFile` has no attachment — it was added to Zotero by DOI lookup (metadata only, no PDF downloaded). The warning is expected; the row will remain `NA` in the audit. `crd_aud_summary()` will flag it as a priority to attach.

---

### Step 3 — Auto-fill quotes

`crd_aud_verify_all()` loads each source document once, then for every unverified row matching that key it scores paragraphs/pages against the `paraphrase` using token overlap. It fills:

- **`quote`** — the best-matching passage from the source
- **`page_or_section`** — paragraph index (docx) or page number (pdf)
- **`verified`** — `"auto"` if score ≥ `min_score`; `"no_match"` if below threshold

```{r verify}
crd_aud_verify_all(
  audit_file = audit_file,
  sources    = sources
)
```

```{r show-verified, echo = FALSE}
d2 <- readr::read_csv(audit_file, show_col_types = FALSE)

# Colour verified column for the table
status_col <- ifelse(
  is.na(d2$verified), "<span style='color:#bbb'>NA</span>",
  dplyr::case_when(
    d2$verified == "auto"     ~ "<span style='background:#FFF2CC;padding:2px 6px;border-radius:3px'>auto</span>",
    d2$verified == "no_match" ~ "<span style='background:#E2EFDA;padding:2px 6px;border-radius:3px'>no_match</span>",
    TRUE ~ d2$verified
  )
)

d2_show <- data.frame(
  citation_key    = d2$citation_key,
  paraphrase      = d2$paraphrase,
  quote           = ifelse(is.na(d2$quote), "", d2$quote),
  loc             = ifelse(is.na(d2$page_or_section), "", d2$page_or_section),
  verified        = status_col,
  stringsAsFactors = FALSE
)

DT::datatable(
  d2_show,
  rownames = FALSE,
  escape   = FALSE,
  colnames = c("citation_key", "paraphrase", "quote", "p/§", "verified"),
  options  = list(
    pageLength = 9, scrollX = TRUE, dom = "t",
    columnDefs = list(
      list(width = "160px", targets = 0),
      list(width = "270px", targets = 1),
      list(width = "270px", targets = 2),
      list(width = "40px",  targets = 3),
      list(width = "80px",  targets = 4)
    )
  ),
  caption = "After crd_aud_verify_all(): auto = matched, no_match = below threshold, NA = no source file."
)
```

**Reading the outcomes:**

- **`auto`** (yellow) — a passage was found in the source scoring above the threshold. The `quote` column shows what the source actually says. Your job: read both, decide if they match.
- **`no_match`** (light green) — the source file exists, but no paragraph scored above the threshold. This happens when the report's phrasing is too different from the source text. It needs manual verification.
- **`NA`** (grey) — no source file is attached in Zotero. Nothing to match against.

Notice row 4: the floodplain growth claim (`"Floodplain rearing habitat supports juvenile growth rates..."`) cites `smith2020SalmonHabitat`, but that PDF covers large woody debris in paragraph 4 — not floodplain rearing. The source simply doesn't contain the claim. This is exactly the kind of mismatch `cred` is designed to surface.

---

### Step 4 — Sort for review

`crd_aud_sort()` rearranges rows so the most actionable are at the top. The `"status"` order is:

`auto` → `no_match` → `corrected` → `no` → `context` → `yes` → `NA`

```{r sort}
crd_aud_sort(audit_file, by = "status")
```

`sort_index` is assigned at write time and never changes, so `crd_aud_sort(by = "report")` always restores the original chapter order.

---

### Step 5 — Summary

`crd_aud_summary()` prints a live status snapshot and — crucially — tells you which PDFs to attach in Zotero to unlock the most NA rows.

```{r summary}
crd_aud_summary(audit_file)
```

In a real report with 200+ citations, the NA sources table is your action list: attach the top-ranked PDF, re-run `crd_aud_verify_all()`, and unlock several rows at once.

---

### Step 6 — Interactive review

`crd_aud_review()` launches a local Shiny app in your browser. It is the recommended interface for working through `auto` rows.

```r
crd_aud_review("background/citation_audit.csv")
```

The app opens filtered to `auto` rows. For each row:

1. Click the row — full `paraphrase` and `quote` appear side by side below the table
2. Read both — does the source support the claim?
3. Set `verified` in the dropdown: `yes`, `no`, or `corrected`
4. Add a `notes` entry if the claim needs correction or context
5. Click **Update row**, move to the next
6. Click **Save to CSV** when done — the plain-text CSV is updated, git history preserved

The app also shows a live progress counter: *N reviewed · N auto · N total*.

---

## The three outcomes explained

| Outcome | What it means | What to do |
|---------|--------------|-----------|
| `auto` | Source file found; passage scored above threshold | Open review app, read quote vs paraphrase, set yes/no/corrected |
| `no_match` | Source file found but no paragraph matched | Check source manually; if claim is valid, set `context` or `yes`; if wrong, set `corrected` |
| `NA` | No source file attached in Zotero | Attach PDF in Zotero → re-run `crd_aud_verify_all()` |

**`no_match` is not an error.** It occurs when:

- The report's phrasing diverges substantially from the source (a sign the claim may have been paraphrased beyond what the source says — worth checking)
- The source provides context rather than a direct quotable finding
- Multiple sources are cited together (`[@a; @b]`) and the paraphrase blends both

---

## Claim type screening

`crd_aud_scr_risk()` populates the `claim_type` column to prioritise your review effort. Statistics (numbers, percentages, years) carry the highest hallucination risk; context citations the lowest.

```{r risk}
d3 <- readr::read_csv(audit_file, show_col_types = FALSE)
d3 <- crd_aud_scr_risk(d3)
table(d3$claim_type)
```

For domain-specific terms, pass an extra pattern:

```r
# Flag species names as statistics for a salmonid report
d3 <- crd_aud_scr_risk(d3, statistic_extra = "chinook|sockeye|coho|pink|chum")
```

Review `statistic` rows first.

---

## Working with your real report

```r
library(cred)

# ── One-time setup ──────────────────────────────────────────────────────────
# Generate the audit CSV from project root (chapter Rmd files only)
crd_aud_write(rmd_dir = ".", out_file = "background/citation_audit.csv")

# ── After adding new Rmd content ────────────────────────────────────────────
# Merge new citations without losing manual edits
crd_aud_upd("background/citation_audit.csv")

# ── Auto-fill from Zotero ───────────────────────────────────────────────────
# Zotero must be open for BBT citation keys; not needed for the SQLite lookup
crd_aud_verify_all("background/citation_audit.csv")

# ── Check progress ──────────────────────────────────────────────────────────
crd_aud_summary("background/citation_audit.csv")

# ── Review session ──────────────────────────────────────────────────────────
# Sort auto rows to the top, open the review app
crd_aud_sort("background/citation_audit.csv", by = "status")
crd_aud_review("background/citation_audit.csv")

# ── Restore report order when done ─────────────────────────────────────────
crd_aud_sort("background/citation_audit.csv", by = "report")
```

### Attaching missing PDFs

`crd_aud_summary()` lists NA sources ranked by claim count — one attachment can unlock many rows. In Zotero:

1. Find the item (use the `citation_key` to search)
2. Right-click → **Add Attachment → Attach File**
3. Re-run `crd_aud_verify_all()` — newly attached files are picked up automatically

### How source matching works

The token scoring approach is simple and fast:

1. Strip `@citekey` markers and markdown punctuation from the `paraphrase`
2. Split into tokens (≥ 4 characters)
3. Score each paragraph/page: proportion of query tokens found in the passage
4. Accept the top-scoring passage if score ≥ `min_score` (default 0.2)

Lower `min_score` to cast a wider net at the cost of weaker matches; raise it to require stricter agreement. The default 0.2 (20% token overlap) works well for specific factual claims with numbers and key terms.
