---
title: "Citation Audit Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citation Audit Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = TRUE
)
```

## The problem

LLM-assisted drafting introduces a specific failure mode: the text is fluent and the citation is real, but the cited source does not actually support the specific claim made. A statistic gets slightly wrong, a finding gets overgeneralised, or a number migrates from one paper to another.

`cred` provides a systematic workflow to catch this. For each citation in a bookdown report, it locates the source document, finds the best-matching passage, and flags claims that need human review.

## Workflow overview

```
Rmd files  →  audit CSV  →  Zotero lookup  →  auto-fill quotes  →  sort  →  review
```

1. **`crd_aud_write()`** — scan Rmd files, extract every `@citekey` with its paraphrase context, write a CSV scaffold.
2. **`crd_zot_src_lookup()`** — query the local Zotero SQLite to find PDF/docx paths for each citation key.
3. **`crd_aud_verify_all()`** — for each source, load it once and fill `quote`, `page_or_section`, and `verified` for all matching rows. (Wraps steps 2 + 4.)
4. **`crd_aud_sort()`** — sort by status (`auto` first) so reviewable rows come first.
5. **`crd_aud_summary()`** — print a status snapshot and attachment priority list.

Open the CSV in Excel or any spreadsheet app. For each `auto` row: read the `quote`, compare to the `paraphrase`, and set `verified` to `yes`, `no`, or `corrected`.

---

## Step-by-step with toy data

`cred` ships toy source files and a minimal Zotero SQLite in `inst/extdata/`:

```
inst/extdata/
├── zotero.sqlite                         # toy Zotero DB (3 items, 2 with attachments)
└── storage/
    ├── TOYPDF1/salmon_habitat.pdf        # PDF: Smith et al. 2020
    └── TOYDOC1/beaver_ecology.docx       # docx: Jones 2019
```

```{r paths}
library(cred)

toy_zotero_dir <- system.file("extdata", package = "cred")
```

### Step 1: Write the audit CSV

`crd_aud_write()` scans all chapter Rmd files (those beginning with four digits) and extracts every citation with its surrounding sentence as `paraphrase`.

```{r write-audit}
# Create two toy Rmd files in a temp directory
rmd_dir <- tempfile("rmd_")
dir.create(rmd_dir)

writeLines(c(
  "# Habitat Quality {#habitat}",
  "",
  "Spawning gravel embeddedness is the primary metric for assessing habitat",
  "quality for chinook salmon [@smith2020SalmonHabitat].",
  "",
  "Riparian vegetation provides critical shade that moderates summer",
  "stream temperatures by 3--7 degrees [@smith2020SalmonHabitat].",
  "",
  "Stream temperature affects the spatial distribution of salmon across",
  "the watershed [@doe2021NoFile]."
), file.path(rmd_dir, "0100-habitat.Rmd"))

writeLines(c(
  "# Beaver Ecology {#beaver}",
  "",
  "Beaver dams increase overwinter survival of juvenile salmonids by",
  "creating slow-water refuges with stable thermal conditions",
  "[@jones2019BeaverEcology].",
  "",
  "Reintroduction of beavers reduces peak flows and extends summer",
  "baseflows [@jones2019BeaverEcology].",
  "",
  "Beaver removal reduced pond habitat extent by 60% between 1850 and",
  "1950 [@jones2019BeaverEcology]."
), file.path(rmd_dir, "0200-beaver.Rmd"))

audit_file <- tempfile("citation_audit_", fileext = ".csv")

crd_aud_write(rmd_dir = rmd_dir, out_file = audit_file)
```

```{r show-audit}
d <- readr::read_csv(audit_file, show_col_types = FALSE)
knitr::kable(d[, c("section", "citation_key", "paraphrase", "verified")])
```

Six rows: three citation keys (`smith2020SalmonHabitat`, `doe2021NoFile`, `jones2019BeaverEcology`), each appearing one to three times.

---

### Step 2: Look up Zotero source paths

`crd_zot_src_lookup()` queries the Zotero SQLite database for the PDF or docx attachment for each key. It uses an immutable read-only URI to avoid locking a running Zotero process.

```{r zotero-lookup}
keys <- unique(d$citation_key)

sources <- crd_zot_src_lookup(keys, zotero_dir = toy_zotero_dir)
print(sources)
```

`doe2021NoFile` has no attachment in this toy database (metadata-only entry), so it is warned and excluded. In a real workflow this is the list of PDFs you need to attach in Zotero before they can be auto-verified.

---

### Step 3: Auto-fill quotes (`crd_aud_verify_all`)

`crd_aud_verify_all()` loads each source document once and, for every unverified row matching that citation key, searches the document using the `paraphrase` as a query. It fills:

- `quote` — best-matching passage
- `page_or_section` — paragraph index (docx) or page number (pdf)
- `verified` — `"auto"` if score ≥ `min_score`; `"no_match"` otherwise

The `sources` argument is optional — if omitted, the function runs `crd_zot_src_lookup()` automatically.

```{r verify-all}
crd_aud_verify_all(
  audit_file = audit_file,
  sources    = sources   # pass pre-looked-up sources to skip re-querying SQLite
)
```

```{r show-filled}
d2 <- readr::read_csv(audit_file, show_col_types = FALSE)
knitr::kable(d2[, c("citation_key", "paraphrase", "quote", "page_or_section", "verified")])
```

Rows citing `doe2021NoFile` remain `NA` because no source file exists. Rows citing `smith2020SalmonHabitat` and `jones2019BeaverEcology` are filled with matching passages.

---

### Step 4: Sort for review

`crd_aud_sort()` rearranges rows to make manual review efficient. The `"status"` sort puts `auto` rows (quotes ready to check) first and `NA` rows (nothing to review) last.

```{r sort}
crd_aud_sort(audit_file, by = "status")
d3 <- readr::read_csv(audit_file, show_col_types = FALSE)
knitr::kable(d3[, c("citation_key", "verified", "sort_index")])
```

`sort_index` is set at write time and never changes — use `crd_aud_sort(by = "report")` to restore original report order at any time.

---

### Step 5: Summary

`crd_aud_summary()` gives a live snapshot of where things stand.

```{r summary}
crd_aud_summary(audit_file)
```

---

## Manual review (the human step)

Open the CSV in Excel, Numbers, or any spreadsheet app. For each `auto` row:

| Column | What to check |
|--------|--------------|
| `paraphrase` | What the report claims |
| `quote` | What the source actually says |
| `page_or_section` | Where in the source to look |
| `verified` | Set to `yes` / `no` / `corrected` |
| `notes` | Record discrepancies; write corrected text |

**`verified` vocabulary:**

| Value | Meaning |
|-------|---------|
| `auto` | Machine-matched — awaiting your review |
| `yes` | Reviewed and confirmed accurate |
| `no` | Claim not supported by source |
| `corrected` | Claim was wrong; you fixed it in the Rmd |
| `no_match` | Source exists but paraphrase did not score above threshold |
| `context` | Citation provides context, not a direct factual claim |
| `NA` | No source file in Zotero — attach a PDF to unblock |

---

## Working with your real report

```r
# One-time: generate the audit CSV (run from project root)
crd_aud_write(rmd_dir = ".", out_file = "background/citation_audit.csv")

# After adding Rmd content: update without losing manual edits
crd_aud_upd("background/citation_audit.csv")

# Auto-fill from Zotero (Zotero must be running for BBT, but not needed for lookup)
crd_aud_verify_all("background/citation_audit.csv")

# Check progress any time
crd_aud_summary("background/citation_audit.csv")

# Sort for review session
crd_aud_sort("background/citation_audit.csv", by = "status")

# Restore report order when done
crd_aud_sort("background/citation_audit.csv", by = "report")
```

### Attaching missing PDFs

`crd_aud_summary()` lists sources with `NA` status ranked by claim count. These are your highest-priority Zotero attachment targets — attaching one PDF can unlock many rows at once.

In Zotero:

1. Find the item (citation key is in the `citation_key` column)
2. Right-click → Add Attachment → Attach File
3. Re-run `crd_aud_verify_all()` — newly attached files are picked up automatically

---

## How source matching works

The token scoring approach is intentionally simple and fast:

1. Strip `@citekey` patterns and markdown punctuation from the `paraphrase`
2. Split into tokens (≥ 4 characters)
3. For each paragraph/page in the source, compute score = proportion of query tokens found
4. Accept the top-scoring passage if score ≥ `min_score` (default 0.2)

This works well for specific factual claims (statistics, named findings). It produces `no_match` for:

- Our own narrative synthesis text (no direct quote exists)
- Highly paraphrased claims where phrasing diverges from source
- Compound `[@a; @b]` citations where the paraphrase is attributed jointly

`no_match` is not an error — it means the row needs manual verification.
